A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-15 tex/codatadp.tex
more

\section{The Decision Procedure} % for (Co)datatypes}
\label{sec:a-ground-decision-procedure-for-co-datatypes}

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Refl}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec, t \teq t
}
\)
\qquad
\rn{Symm}
\(
\inferrule{
 t_1 \teq t_2 \in \Ec
}{
 \Ec := \Ec, t_2 \teq t_1
}
\)
\qquad
\rn{Conflict}
\(
\inferrule{
  t_1 \teq t_2, t_1 \tneq t_2 \in \Ec
}{
  \bot
}
\)
\\[3.7ex]
\rn{Trans}
\(
\inferrule{
  t_1 \teq t_2, t_2 \teq t_3 \in \Ec
}{
  \Ec := \Ec, t_1 \teq t_3
}
\)
\qquad
\rn{Cong} 
\(
\inferrule{
  \vec t \teq \vec u \in \Ec \quad f( \vec t ), f( \vec u ) \in \tEc
}{
  \Ec := \Ec, f( \vec t ) \teq f( \vec u )
}
\)
\\[3.7ex]
\rn{Unify$_1$} 
\(
\inferrule{
  \const{C}_1( \vec t ) \teq \const{C}_1( \vec u ) \in \Ec
}{
  \Ec := \Ec, \vec t \teq \vec u
}
\)
\qquad
\rn{Unify$_2$} 
\(
\inferrule{
  \const{C}_1( \vec t ) \teq \const{C}_2( \vec u ) \in \Ec
}{
  \bot
}
\)
\end{tabular}
\caption{Rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Cyclic}
\(
\inferrule{
  t_1 : \tau
  \quad
  \tau \in \Data
  \quad
  %\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset
  \Val \ec{t_1} = \mu x. t_2
  \quad
  x \in FV( t_2 )
}{
  \bot
}
\)
\\[3.7ex]
\rn{Bisimilar}
\(
\inferrule{
 t_1, t_2 : \tau
 \quad
 \tau \in \Codata
 \quad
 %\tpath{\Ec}{\ec{t_1}} = \tpath{\Ec}{\ec{t_2}}\sigma \neq \emptyset
 \Val \ec{t_1} \vsim \Val \ec{t_2}
}{
 \Ec := \Ec, t_1 \teq t_2
}
\)
\end{tabular}
\caption{Rules for cyclicity and bisimilarity%.
}
\label{fig:ab-rules}
\end{figure}

@@

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve a mapping $\Val$ from equivalence classes to terms containing $\mu$-bindings.
Recall that $\mu$-bindings can be used for representing (possibly cyclic) terms and values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure either of the form
$\mu x. \const{C}( \vec t )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
For consistency, we will also consider $\mu x. \const{c}$ to be a $\mu$-term
in the case where $\const{c}$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$
and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if and only if it contains no free variables.
A $\mu$-term is \emph{cyclic} if and only if it contains a bound variable.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables,
in which case we will commonly say $t$ and $s$ are \emph{bisimilar}.
For instance, 
$\mu x. \const{C}( x ) \vsim \mu y. \const{C}( y )$ and
$\mu x, \const{D}( y, x ) \vsim \mu z. \const{D}( y, z )$, but
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{C}( x )$,
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{D}( z, x )$ and
$\mu x. \const{D}( y, x ) \not\vsim \mu y. \const{D}( y, x )$.
Also, note that
two $\mu$-terms may represent the same value while being not bisimilar, as in
$\mu x. \const{C}( \mu y. \const{C}( x ) ) \not\vsim \mu x. \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Singleton}
\(
\inferrule{
  t, u : \tau \in \tEc
  \quad
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\\[3.7ex]
\rn{Split} 
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\end{tabular}
\caption{Rules for branching%.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

@@

\begin{definition}[Normal Model] \label{def:norm-model}
A model $\M$ is \emph{normal} if and only if:
\begin{enumerate}
\item 
for all types $\tau$, 
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are 
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$,
\item 
for all constructor terms $\const{C}( \vec t )$ of type $\tau$,
$\M( \const{C}( \vec t ) )$ is the value 
in $\M( \tau )$ that is a renaming of 
$\nf{(\mu x. \const{C}( \M( \vec t ) ))}$ where $x$ is fresh, and
\item 
for all selector terms $s^j_i( t )$ of type $\tau$,% for some $\vec u$,
$\M( s^j_i( t ) )$ is the value
in $\M( \tau )$ that is a renaming of 
$\M( t )\interp{u_j}$
when $\M( t )$ is of the form $\mu x. \const{C}_i( \vec u )$.
\end{enumerate}
\end{definition}

@@

\section{Implementation}
\label{sec:the-theory-solver}

@@

\section{Experimental Results}
\label{sec:experimental-results}

>>>>

\section{The Ground Decision Procedure} % for (Co)datatypes}
\label{sec:a-ground-decision-procedure-for-co-datatypes}

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec,\: t \teq t
}
\)
\rn{Refl}
\qquad
\(
\inferrule{
 t_1 \teq t_2 \in \Ec
}{
 \Ec := \Ec,\: t_2 \teq t_1
}
\)
\rn{Symm}
\qquad
\(
\inferrule{
  t_1 \teq t_2,\; t_1 \tneq t_2 \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\\[5\jot]
\(
\inferrule{
  t_1 \teq t_2,\; t_2 \teq t_3 \in \Ec
}{
  \Ec := \Ec,\: t_1 \teq t_3
}
\)
\rn{Trans}
\qquad
\(
\inferrule{
  \vec t \teq \vec u \in \Ec \quad \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \Ec := \Ec,\: \const f( \vec t \,) \teq \const f( \vec u )
}
\)
\rn{Cong} 
\\[5\jot]
\(
\inferrule{
  \const{C}_1( \vec t \,) \teq \const{C}_1( \vec u ) \in \Ec
}{
  \Ec := \Ec,\: \vec t \teq \vec u
}
\)
\rn{Unify$_1$} 
\qquad
\(
\inferrule{
  \const{C}_1( \vec t \,) \teq \const{C}_2( \vec u ) \in \Ec
}{
  \bot
}
\)
\rn{Unify$_2$} 
\end{tabular}
\caption{Rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\(
\inferrule{
  \tau \in \Data
  \quad
  t_1 : \tau
  \quad
  %\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset
  \Val \ec{t_1} = (\vthinspace\mu x.\; t_2)
  \quad
  x \in FV( t_2 )
}{
  \bot
}
\)
\rn{Cyclic}
\\[5\jot]
\(
\inferrule{
 \tau \in \Codata
 \quad
 t_1, t_2 : \tau
 \quad
 %\tpath{\Ec}{\ec{t_1}} = \tpath{\Ec}{\ec{t_2}}\sigma \neq \emptyset
 \Val \ec{t_1} \vsim \Val \ec{t_2}
}{
 \Ec := \Ec,\: t_1 \teq t_2
}
\)
\rn{Bisimilar}
\end{tabular}
\caption{Rules for cyclicity and bisimilarity%.
}
\label{fig:ab-rules}
\end{figure}

@@

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve a mapping $\Val$ from equivalence classes to terms containing $\mu$-bindings.
Recall that $\mu$-bindings can be used for representing (possibly cyclic) terms and values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure either of the form
$\mu x. \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
For consistency, we will also consider $\mu x. \const{c}$ to be a $\mu$-term
in the case where $\const{c}$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$
and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if and only if it contains no free variables.
A $\mu$-term is \emph{cyclic} if and only if it contains a bound variable.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables,
in which case we will commonly say $t$ and $s$ are \emph{bisimilar}.
For instance, 
$\mu x. \const{C}( x ) \vsim \mu y. \const{C}( y )$ and
$\mu x, \const{D}( y, x ) \vsim \mu z. \const{D}( y, z )$, but
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{C}( x )$,
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{D}( z, x )$ and
$\mu x. \const{D}( y, x ) \not\vsim \mu y. \const{D}( y, x )$.
Also, note that
two $\mu$-terms may represent the same value while being not bisimilar, as in
$\mu x. \const{C}( \mu y. \const{C}( x ) ) \not\vsim \mu x. \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\(
\inferrule{
  t, u : \tau \in \tEc
  \quad
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\rn{Split} 
\end{tabular}
\caption{Rules for branching%.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

@@

\begin{definition}[Normal Model] \label{def:norm-model}
A model $\M$ is \emph{normal} if and only if:
\begin{enumerate}
\item 
for all types $\tau$, 
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are 
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$,
\item 
for all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value 
in $\M( \tau )$ that is a renaming of 
$\nf{(\mu x. \const{C}( \M( \vec t \,) ))}$ where $x$ is fresh, and
\item 
for all selector terms $s^j_i( t )$ of type $\tau$,% for some $\vec u$,
$\M( s^j_i( t ) )$ is the value
in $\M( \tau )$ that is a renaming of 
$\M( t )\interp{u_j}$
when $\M( t )$ is of the form $\mu x. \const{C}_i( \vec u )$.
\end{enumerate}
\end{definition}

@@

\section{Implementation as a Theory Solver in CVC4}
\label{sec:the-theory-solver}

@@

\section{Evaluation on Isabelle Problems}
\label{sec:experimental-results}
